---
title: "Lab 4 answer key"
author: "Kristi Gdanetz MacCready"
date: "9/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we will start working with some of the other measurements associated with the microbiome samples; breath gases, pH, and bristol stool scores. 

```{r}
# Import the file
scfa_wkly <- read_delim("~/Box Sync/Schmidt_Lab/Bio201_teaching/Protocol_Rmd_docs/Lab5/raw_data/SCFA_wkly_v05.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
```

### Homework 4.1 
Apply the group and summarise functions to determine the mean concentration of all three SCFAs for each participant, during each week of the study. Can you do this without creating any intermediate data frames? 

```{r}
# write code here
scfa_wkly <- scfa_data %>%
  group_by(participant_id, study_week, semester, supplement_consumed) %>%
  summarize(mean_acetate = mean(acetate_mmol_kg, na.rm = TRUE), 
            mean_butyrate = mean(butyrate_mmol_kg, na.rm = TRUE), 
            mean_propionate = mean(propionate_mmol_kg, na.rm = TRUE))
```

### Spread 
### Homework 4.2
Code for each SCFA
```{r}
butyrate_wide <- scfa_wkly %>%
  select(participant_id, semester, supplement_consumed, study_week, butyrate_mean) %>% #drop extra columns
  filter(study_week == "week1" | study_week == "week3") %>% #drop extra weeks
  spread(study_week, butyrate_mean) %>% #convert to wide format
  na.omit(week1, week3) %>% #drop NAs from select columns
  mutate(delta_butyrate = week3 - week1) %>% #calculate difference 
  rename(but_wk1 = week1,
         but_wk3 = week3)

head(butyrate_wide)
```

Create a wide data frame for the other SCFAs.
```{r}
# copy+paste and edit code chunk for acetate
acetate_wide <- scfa_wkly %>%
  select(participant_id, semester, supplement_consumed, study_week, acetate_mean) %>% #drop extra columns
  filter(study_week == "week1" | study_week == "week3") %>% #drop extra weeks
  spread(study_week, acetate_mean) %>% #convert to wide format
  na.omit(week1, week3) %>% #drop NAs from select columns
  mutate(delta_acetate = week3 - week1) %>% #calculate difference 
  rename(ace_wk1 = week1,
         ace_wk3 = week3)

head(acetate_wide)
```

```{r}
# copy+paste and edit code chunk for propionate
propionate_wide <- scfa_wkly %>%
  select(participant_id, semester, supplement_consumed, study_week, propionate_mean) %>% #drop extra columns
  filter(study_week == "week1" | study_week == "week3") %>% #drop extra weeks
  spread(study_week, propionate_mean) %>% #convert to wide format
  na.omit(week1, week3) %>% #drop NAs from select columns
  mutate(delta_propionate = week3 - week1) %>% #calculate difference 
   rename(pro_wk1 = week1,
         pro_wk3 = week3)

head(propionate_wide)
```

### Homework 4.3
We are interested in the average difference of fecal butyrate by supplement type, how would you do this? 
```{r}
# write code here
but_diff <- butyrate_wide %>%
  group_by(supplement_consumed) %>%
  summarise(delta_mean_but = mean(delta_butyrate, na.rm = TRUE))
```

Repeat this process for each SCFA. How would you plot your results? 
```{r}
# copy+paste and edit code chunk for acetate
ace_diff <- acetate_wide %>%
  group_by(supplement_consumed) %>%
  summarise(delta_mean_ace = mean(delta_acetate, na.rm = TRUE))
```

```{r}
# copy+paste and edit code chunk for propionate
prop_diff <- propionate_wide %>%
  group_by(supplement_consumed) %>%
  summarise(delta_mean_pro = mean(delta_propionate, na.rm = TRUE))
```

```{r}
# copy+paste and edit code chunks for plots
ggplot(but_diff, aes(x = supplement_consumed,
                     y = delta_mean_but)) +
  geom_jitter()
```

These plots are fine, however they could be graphed more efficiently with facets. This introduces a new problem: combining data frames.

### Join
### Homework 4.4 
Part 1: Join and update the column names of the three SCFA wide data frames, calculate the weekly total SCFA for each participant, calculate the difference in total SCFA across weeks.
```{r}
total_scfas <- inner_join(acetate_wide, butyrate_wide,
                   by = c("participant_id", "semester", "supplement_consumed")) %>%
  full_join(propionate_wide, 
            by = c("participant_id", "semester", "supplement_consumed")) %>%
  mutate(total_wk1 = ace_wk1 + but_wk1 + pro_wk1,
         total_wk3 = ace_wk3 + but_wk3 + pro_wk3,
         delta_total = total_wk3 - total_wk1)
```

Part 2: Plot the change in total SCFA for each semester, faceting for each supplement consumed. Clean up the axes and labels. Extra challenge: Can you use pipes to execute Parts 1 and 2 as one long command?
```{r}
ggplot(data = total_scfas,
       aes(x = supplement_consumed, y = delta_total, color = semester)) +
  geom_point() +
  facet_grid(~semester, scales = "free")
```


### Gather
### Homework 4.5 
```{r}
# copy+paste and edit code here to create but_long
but_long <- total_scfas %>%
  select(-starts_with("delta_"), -starts_with("ace_"), -starts_with("pro_"), -starts_with("total")) %>% 
  rename(week1 = but_wk1,
         week3 = but_wk3) %>% 
  gather(key = study_week, value = butyrate, week1, week3)
```

```{r}
# copy+paste and edit code here to create pro_long
pro_long <- total_scfas %>%
  #drop diff columns & extra SCFAs
  select(-starts_with("delta_"), -starts_with("ace_"), -starts_with("but_"), -starts_with("total")) %>% 
  rename(week1 = pro_wk1,
         week3 = pro_wk3) %>% #convert weeks to consistent format for join
  gather(key = study_week, value = propionate, week1, week3)
```

Check the dimensions of the three long data frames you just created, select the appropriate type of join, and create a new data frame called scfa_long. 
```{r}
scfa_long <- full_join(ace_long, but_long,
                       by = c("participant_id", "semester", "supplement_consumed", "study_week")) %>%
  full_join(pro_long, by = c("participant_id", "semester", "supplement_consumed", "study_week"))
```

Now that the data are in long format, calculate the weekly mean for each SCFA per participant. Make use of the code from the section above, just swap out the variables for the data frames and column names. 
```{r}
# copy+paste and edit code here
wkly_means <- scfa_long %>%
  group_by(participant_id, semester, supplement_consumed, study_week, scfa) %>%
  summarise(mean_scfa= mean(concentration_mmol_kg))
```


### Export
```{r}
write_delim(total_scfas, path = "~/Box Sync/Schmidt_Lab/Bio201_teaching/Protocol_Rmd_docs/Lab5/raw_data/scfa_wide.txt",
            delim = "\t", na = "NA", col_names = TRUE)
```

### Homework 4.6

