---
title: 'Lab 4: Graphics part 2'
author: "Kristi Gdanetz MacCready"
date: "9/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
set.seed(7)
```

# Import data
If these data are not already loaded in the global environment, import them now. Two of the datasets we will use for this section are the cleaned up data we generated last week. 
```{r}
#scfa_wide <- read_delim(file = "Lab4/curated_data/scfa_wide.txt", delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA"))

pH_BSS_wkly <- read_delim(file = "Lab4/curated_data/pH_BSS_wkly.txt",
           delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA"))

scfa_indv_qc <- read_delim(file = "Lab3/curated_data/scfa_indv_qc.txt", 
           delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA"))

SCFA_wkly_qc <- read_delim(file = "Lab3/curated_data/SCFA_wkly_qc.txt", 
           delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA"))
```


# Intro

As mentioned in Lab 2, one of the over arching questions for this course is: "Does addition of a fiber supplement increase microbial fermentation in the gut?" We are measuring microbial fermentation today with butyrate concentration in feces. Today and last week we cleaned up these data and determined the average concentration of each indivdual during the study weeks.

For the second module on graphing we will use the scfa_wkly_qc data frame to illustrate the appropriate types of plots for the families of data we will encounter in this course. 

# Graph 1 continuous and 1 categorical (discrete variable)

We are going to build upon the graphing skills learned last week to learn some additional types of plots, and how to choose the best plot type for your dataset. In this lesson we are going to use the term categorical, however these are called discrete on the ggplot cheatsheet. 

### Bar plots

Bar plots are a popular way to display continuous data that has been divided into categories (one discrete + one continuous variables). However they can represent the data in a misleading format because they do not reflect all the variability within the discrete variable(s). For example, plot the mean SCFA concentration for each supplement group, or study week. Instead of using geom_point, we will use geom_col to build bar plots.

```{r}
ggplot(data=scfa_indv_qc, aes(x=study_week, y=butyrate_mmol_kg)) +
  geom_col(position="stack") 
```

Notice the y-axis. We know a butyrate concentration of 20,000 is not correct. Here geom_col is adding the values from all individuals in a group (study week). What we really want is some central measure for each group. So to build a plot that is more accurate, we first must calculate the mean value for each group. We will calculate means for each race/ethnicity and week, because an average of butyrate concentration between weeks 1 and 3 doesn't really make sense.

```{r}
# calculate weekly mean per participant
scfa_wkly_qc <- scfa_indv_qc %>%
  group_by(participant_id, study_week, race_ethnicity, semester, supplement_consumed) %>%
  summarize(mean_wkly_butyrate = mean(butyrate_mmol_kg, na.rm=TRUE), 
            sd_wkly_butyrate = sd(butyrate_mmol_kg, na.rm=TRUE))
```
Inspect the results. We have some individuals where race/ethnicity was not recorded, should these be included?
```{r}
scfa_wkly_qc <- scfa_indv_qc %>%
  drop_na(race_ethnicity) %>%
  group_by(participant_id, study_week, race_ethnicity, semester, supplement_consumed) %>%
  summarize(mean_wkly_butyrate = mean(butyrate_mmol_kg, na.rm=TRUE), 
            sd_wkly_butyrate = sd(butyrate_mmol_kg, na.rm=TRUE))
```

```{r}
# calculate means per race/ethnicity 
scfa_wkly_qc2 <- scfa_wkly_qc %>%
  group_by(study_week, race_ethnicity) %>%
  summarize(mean_butyrate = mean(mean_wkly_butyrate, na.rm=TRUE), 
            sd_butyrate = sd(mean_wkly_butyrate, na.rm=TRUE))
```
Inspect the results. We have 6 unique instances of "2 or more...". Should these be grouped together? There are also combinations of study week + race/ethnicity where it was not possible to calculate a standard deviation due to a small sample size. 

```{r}
ggplot(data=scfa_wkly_qc2, aes(x=study_week, y=mean_butyrate)) +
  geom_col(position="stack") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5))
```

This is better, but still isn't quite right. There is nothing in this plot distinguising between race/ethnicity for the geom_col function, we can modify x (position on x-axis), y (position on the y-axis), fill (color of the bar itself), alpha (transparency of the fill color), color (color of the line that surrounds each bar), linetype (amount of hashing of the line that surrounds each bar), and size (thickness of the surrounding line) aesthetics. We will use these aesthetics in addition to setting geom_col parameters to obtain a more informative figure.  

```{r}
ggplot(data=scfa_wkly_qc2, aes(x=study_week, y=mean_butyrate, fill=race_ethnicity)) +
  geom_col(position="stack") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5))
```
You will likely have to adjust your window to view the plot because some of the group names are very long. 

Stacked bars are impossible to compare across groups. We can set the bars side-by-side instead, and do some other cleaning-up of the plot. We’ll map our fill aesthetic to the “race/ethnicity” variable, and we’ll add a position="dodge" argument to geom_col. The “dodge” will bump the bars to the side so they don’t stack on top of each other. To update the y-axis label and placement of the tick marks; use the scale_y_continuous function. The xlab function is used to hide the label. 

```{r}
ggplot(data=scfa_wkly_qc2, aes(x=study_week, y=mean_butyrate, fill=race_ethnicity)) +
  geom_col(position="dodge") +
  scale_y_continuous(name="Mean fecal butyrate (mmol/kg)",
                     limits = c(0, 25),
                     breaks = c(0, 5, 10, 15, 20, 25)) +
  xlab(NULL) 
```

There are obviously some issues with our groups here, we have 6 combinations of "2 or more...", some groups do not appear to have measurements for all weeks of the study, and there are "Other" and "NA" groups. We will come back to these issues when we discuss factors later today. 

One last thing with bar plots, you should represent the variation in the data by adding error bars. These are added with the function geom_errorbar(). This geom will take x, y, ymax, ymin, alpha, color, linetype, size, width as aesthetics. The two new aesthetics here are ymax and ymin, which refer to where the top and bottom lines should be drawn on the error bars, respectively. 


```{r}
# create the plot 
ggplot(scfa_wkly_qc2, aes(x=study_week, 
                          y=mean_butyrate, 
                          fill=race_ethnicity)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymax = mean_butyrate + sd_butyrate, 
                    ymin = mean_butyrate - sd_butyrate), 
                width=0.5, 
                position = "dodge") +
   scale_y_continuous(name="Mean fecal butyrate (mmol/kg)",
                     limits = c(0, 25),
                     breaks = c(0, 5, 10, 15, 20, 25)) +
  xlab(NULL)  
```

What a mess. There are two problems with this result, one far more prominent than the other. First, the standard deviations are huge; this is one of the main reasons bar plots are an undesirable tool for presenting mean data. The second problem is the error bars do not align with the groups. 

This example helps to illustrate the problems with bar plots. In the next section we will visualize the same data in a few other ways to represent continuous data from multiple categories, to illustrate the ideal methods to represent data. Much of the syntax we used with geom_col is the same for these other geoms.

### Jitter plots

As mentioned above, bar plots obscure the data making it hard to determine whether the data are normally distributed and the number of observations. An alternative to this is a strip or jitter plot where the y-axis values are plotted for each observation in the category. These plots are generated with geom_jitter.

Notice that this requires the full data frame (weekly means for each participant), not the dataframe with the summary statistics.
```{r}
ggplot(scfa_wkly_qc, aes(x=study_week, 
                         y=mean_wkly_butyrate, 
                         color=race_ethnicity)) +
  geom_jitter(shape=1, 
              size=2)  + 
  #geom_errorbar(aes(ymax=mean_butyrate+sd_butyrate, ymin=mean_butyrate-sd_butyrate), width=0.5) +
  #scale_fill_manual() +
  #scale_x_discrete(name = " ", breaks=c(), labels=c()) +
  ylab("Mean butyrate (mmol/kg)") +
  xlab(NULL)
```

Notice the points in each category are close together, to alter the jitter along the x-axis, use the width argument.

```{r}
ggplot(scfa_wkly_qc, aes(x=study_week, 
                         y=mean_wkly_butyrate, 
                         color=race_ethnicity)) +
  geom_jitter(shape=1, 
              size=2, 
              width=0.2)  + 
  ylab("Mean butyrate (mmol/kg)") +
  xlab(NULL)
```

This format makes it difficult to identify patterns in the data. Another option is to use the position_jitterdodge function with the position argument in geom_jitter. Recall that “dodge” means move the points so they don’t overlap and “jitter” randomizes the x-position of the points.

It looks like our race/ethnicity groups are still overlapping a bit. We can give a jitter.width and dodge.width value to position_jitterdodge to eliminate that overlap.

```{r}
ggplot(scfa_wkly_qc, aes(x=study_week, 
                         y=mean_wkly_butyrate, 
                         color=race_ethnicity)) +
  geom_jitter(shape=1, 
              size=2,  
              position = position_jitterdodge(jitter.width = 0.1, 
                                              dodge.width = 1))  + 
  ylab("Mean butyrate (mmol/kg)") +
  xlab(NULL)
```

Jitter plots are useful because they present all of the data. However, these can get a bit messy when there are a large number of observations (like the plots we generated in Lab 3). Additionally, there is no identification of the mean, median, or ranges. However, we can use geom_errorbar to add these features, but without addressing our uneven groups, that won't achieve a better result than when we used geom_col. 


### Box & violin plots

An alternative to the jitter plot is the box plot. Below, generate a box plot using the geom_boxplot function in much the same way we did earlier with the geom_jitter. However, a box plot may not be meaningful if there aren’t many observations. 

Box plots have a traditional definition for what the elements represent, however some researchers modify this and may not clearly disclose what the elements represent. Usually, the line through the middle of the rectangle is the median value and the lower and upper edges of the rectangle represent the 25th and 75th percentiles. The whiskers extend to the largest value greater than 1.5 times the difference between the 25th and 75th percentiles; it’s a way to represent outliers. 


```{r}
# same code swapped geom function
ggplot(scfa_wkly_qc, aes(x=study_week, 
                         y=mean_wkly_butyrate, 
                         color=race_ethnicity)) +
  geom_box()  + 
  ylab("Mean butyrate (mmol/kg)") +
  xlab(NULL)
```

```{r}
# overlay jitter and box
ggplot(scfa_wkly_qc, aes(x=study_week, 
                         y=mean_wkly_butyrate, 
                         color=race_ethnicity)) +
  geom_box(outlier.shape = NA) + #use outlier arg to deal with duplicate data points
  geom_jitter(shape=1, 
              size=2,  
              position = position_jitterdodge(jitter.width = 0.1, 
                                              dodge.width = 1))  + 

  ylab("Mean butyrate (mmol/kg)") +
  xlab(NULL)
```

### Homework 4.5
```{r}
#<Adjust shape and transparency>
```

A downside of box plots is they indicate the positions of the median values, where the statistical tests will we use next week determine differences between groups based on mean values. 

You *could* alter the values of the box plots to indicate the mean, however, that would likely be misleading to anyone viewing the plot since means are not typically used. It is more desirable to use another plot type instead.

In the last box plot example, we plotted the data points on top of the box plot. Depending on the number of data points, this can become cluttered. An alternative is the violin plot, where the position along the left axis indicates the density of values at that position on the y-axis. You can create violin plots very much in the same way as strip carts and box plots using the geom_violin. 

```{r}
#<add example, geom_violin()>
```

Another alternative method is to manually add annotations of means and standard deviations. 


# Figure Annotations
As mentioned above, we can manually add lines for means and standard deviations. When plotting a difference value, it is often useful to draw a line at zero, this helps viewers quickly distinguish positive and negative values. Later in the semester, you may want to add significance annotations to figures. This section will introduce a handful of functions that will be useful to accomplish these tasks. 

### Homework 4.6 
Copy the code from the section on joins, which plots the difference in each SCFA. Notice the 0 centered on the y-axis. To add a horizontal line at the zero position, add the function for geom_hline() to the end of your code block. 

Remember that each geom layer is added in order, so if you add geom_hline() at the end, the line annotation is actually on top of our data points. Move the geom_hline command so it occurs before geom_jitter, now all points are laid over this line, and the line is no longer obscuring these data. 
```{r}
#Set yintercept=0 and linetype="dashed"
```

To add lines for mean, standard deviation, or any other summary statistics we have to calculate these values before generating the figure. Calculate the number of samples, the mean, the standard deviations, and the standard error (StdErr = StdDev/(sqrt(N))) based on figure groups using the code below:

Assign each of these values to an object, then we will call on these values using geom_errorbar. We will use the geom_errorbar command consecutively, once to draw values for mean and a second time to draw values for the standard error.  

You may be wondering, which should type of error should be included in my plot, standard deviation or standard error? 

* Standard deviation is use to indicate the variation around the mean. Useful to indicate variation in measurements. 
* Standard error accounts for the variance (via standard deviation) and the sample size. Good to indicate uncertaintiy in mean estimate. 

I prefer to include the standard deviation in my figures, becuase it is a better representation of the biologial variation of the sample population. You can include either when creating figures for your assignments in this class, just be sure to state which in the figure descriptions.

Assigning a variable to the x aesthetic, matches the value to categorical variable. Assignment to the y aesthetic provides the location along the y-axis (a position along the continuous scale). The ymin and ymax aesthetics indicate the lower and upper bounds of the error bars. When we use geom_errobar to plot a mean, the ymin and ymax values are the same. 
```{r}

```

### Homework 4.7 
As we observed during the bar plot section of the analysis, the default width for the error bars could use some modifications. I prefer to plot the mean the same width as the jitter, and the standard deviations around ½ to ¼ width. Add a width argument (outside of the aesthetics argument) to update the errorbar width to something reasonable.
```{r}

```


# Factors

Whenever we’ve grouped our data by semester, the groups are ordered alphabetically within each group (i.e., fall15, fall16, ..., winter19) rather than in order by time (winter15, fall15, winter16, ..., winter19). To fix the ordering, we need to cast these variables as factors. Factors can be a troublesome feature within R for users, however there are packages within the tidyverse which make working with factors much easier. 

Factors are a special type of data for representing categorical data. Characters are another type of data for representing categorical data, but the categories are ordered alphabetically. However as described above we want to order them by time, so we need to convert the characters into factors. We can also rename factors so that "winter15" is displayed as "Winter 2015", or if we are pressed for space we could use "W15". This is a similar idea as when we relabeled plot axes or recoded incorrect variables.
 
Let’s return to the example of... 

Copy the code which plots the difference in SCFA concentrations. We will simplify this a little to illustrate facets, modify to remove all SCFAs except butyrate. Remove any horizontal annotations, make sure the x aesthetic is set to semester.

```{r}

```

Observe the default behavior (semester is a character data type). The easiest way to deal with factors is to set their order during the curation and processing of raw data before it is saved to the curated data folder.
```{r}

```

In a normal workflow the code chunk above could be added to the end of the data curation pipeline. If you want to convert data to a factor using pipes, use the factor() function within mutate():

```{r}

```

Take the new data frame and re-run the code to generate the figure, observe the updated order of the semesters. 
```{r}

```

# Graph 2 continuous variables

Plotting two continutous variables, each axis is represented by a range of numbers. Each data point has an x and y measure. (You may know these types of plots as scatter plots from Excel). These are acheived in R using geom_point(). These are useful to visualize correlations, for example abundance of a fiber degrader and SCFA concentration. 

```{r}

```


# Graph 2 categorical variables
These are tricky to visualize, and don't have clear applications. If you plot supplement consumed vs. semester, all you can really conclude is the number of particpants in each combination of variables. Personally I think these type of data are beter represented in a table. 

```{r}

```




