---
title: "Lab5 Answers"
author: "Kristi Gdanetz MacCready"
date: "10/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

# Load packages

```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
set.seed(7)
```



# Homework 5.1
Repeat the process to conduct the paired t-test for the 1xdaily group. Remember to check the assumptions. Write your conclusions regarding the test as a comment at the end of the code block. 
```{r}
# copy + paste code, and update column and dataframes names to run test

# Check assumptions for each week of the 1xdaily groups 
wk1_1x <- butyrate_long %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk1_1x$butyrate_mean) 
ggplot(wk1_1x, aes(x = butyrate_mean)) + geom_histogram()

wk3_1x <- butyrate_long %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk3_1x$butyrate_mean) 
ggplot(wk3_1x, aes(x = butyrate_mean)) + geom_histogram()

# join these data frames back together
x1_df <- inner_join(x = wk1_1x, y = wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(butyrate_mean_wk1 = butyrate_mean.x,
         butyrate_mean_wk3 = butyrate_mean.y) %>%
  select(-starts_with("study_week"))

# code to run paired t-test 
t.test(x = x1_df$butyrate_mean_wk1, y = x1_df$butyrate_mean_wk3, 
       alternative = "less", paired = TRUE) #p-value = 0.2367

wilcox.test(x = x1_df$butyrate_mean_wk1, 
            y = x1_df$butyrate_mean_wk3, 
            alternative = "less", paired = TRUE) #p-value = 0.2269
```

# Homework 5.2
Generate plots to determine if methane, acetate, propionate increased during week 3 of the study for inidivduals who consumed BRMPS. Generate plot to determine if pH decreased during week 3 of the study. 

### Methane
```{r}
# import data 
breath_wkly <- read_delim("~/Box Sync/Schmidt_Lab/Bio201_teaching/Protocol_Rmd_docs/Lab5/curated_data/breath_wkly.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))

# plot
breath_wkly %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  ggplot(aes(x = study_week, y = ch4_mean, color = study_week)) + 
  geom_violin() + geom_jitter() + 
  facet_grid(~frequency) + 
  xlab(NULL) + 
  ylab("Breath methane (ppm)") + 
  theme(legend.position = "none")
```

### Acetate
```{r}
scfa_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = acetate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + geom_jitter() + 
  facet_grid(~frequency) + 
  xlab(NULL) + 
  ylab("Acetate (mmol/kg)") + 
  theme(legend.position = "none")

scfa_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = acetate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + geom_jitter() + 
  xlab(NULL) + 
  ylab("Acetate (mmol/kg)") + 
  theme(legend.position = "none")
```

### Propionate
```{r}
scfa_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = propionate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + geom_jitter() + 
  facet_grid(~frequency) + 
  xlab(NULL) + 
  ylab("Propionate (mmol/kg)") + 
  theme(legend.position = "none")

scfa_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = propionate_mean, 
             color = study_week)) + 
  geom_violin() + geom_jitter() + 
  xlab(NULL) + 
  ylab("Propionate (mmol/kg)") + 
  theme(legend.position = "none")
```

### pH 
```{r}
# import data 
pH_wkly <- read_delim("~/Box Sync/Schmidt_Lab/Bio201_teaching/Protocol_Rmd_docs/Lab5/curated_data/pH_wkly.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
# plot
pH_wkly %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + geom_jitter() + 
  facet_grid(~frequency) + 
  xlab(NULL) + 
  ylab("pH") + 
  theme(legend.position = "none")

pH_wkly %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean, 
             color = study_week)) + 
  geom_violin() + geom_jitter() + 
  xlab(NULL) + 
  ylab("pH") + 
  theme(legend.position = "none")
```

# Homework 5.3
Check assumptions on data sets for individuals who consumed BRMPS: breath methane, pH, acetate, propionate
### Methane
```{r}
# check sample size
breath_wkly %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>% 
  group_by(frequency, study_week) %>% 
  summarise(sample_size = n()) 

# normal distribution 
ch4_df <- breath_wkly %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS")
shapiro.test(ch4_df$ch4_mean) #p-value < 2.2e-16
ggplot(ch4_df, aes(x=ch4_mean)) +
  geom_histogram() 
qqnorm(ch4_df$ch4_mean); qqline(ch4_df$ch4_mean)

# variances
# code to extract first group (1xdaily)
once_grp <- ch4_df %>%
  filter(study_week == "week3",
         frequency == "1xdaily") 
# code to extract second group (2xdaily)
twice_grp <- ch4_df %>%
  filter(study_week == "week3",
         frequency == "2xdaily") 
var.test(x = once_grp$ch4_mean, 
         y = twice_grp$ch4_mean, 
         alternative = "two.sided") #p-value = 0.001711

# non-normal, unequal variances, use wilcoxon test 
```

### Acetate
```{r}
# sample size
scfa_long %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>% 
  group_by(frequency, study_week) %>% 
  summarise(sample_size = n()) 

# normality
ace_df <- scfa_long %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(-starts_with("but"), -starts_with("pro"))

shapiro.test(ace_df$acetate_mean) #p-value = 4.481e-11
ggplot(ace_df, aes(x=acetate_mean)) +
  geom_histogram() #histograms show the number of data points (count) at each value
qqnorm(ace_df$acetate_mean); qqline(ace_df$acetate_mean)

# variances
once_grp <- scfa_long %>%
  filter(study_week == "week3",
         frequency == "1xdaily") 
twice_grp <- scfa_long %>%
  filter(study_week == "week3",
         frequency == "2xdaily") 
var.test(x = once_grp$acetate_mean, 
         y = twice_grp$acetate_mean, 
         alternative = "two.sided") #p-value = 0.1585

# non-normal, equal variances
```

### Propionate
```{r}
# sample size
scfa_long %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>% 
  group_by(frequency, study_week) %>% 
  summarise(sample_size = n()) 

# normality
pro_df <- scfa_long %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(-starts_with("ace"), -starts_with("but"))
shapiro.test(pro_df$propionate_mean) #p-value = 2.014e-12
ggplot(pro_df, aes(x=propionate_mean)) +
  geom_histogram() #histograms show the number of data points (count) at each value
qqnorm(pro_df$propionate_mean); qqline(pro_df$propionate_mean)

# variances
once_grp <- scfa_long %>%
  filter(study_week == "week3",
         frequency == "1xdaily") 
twice_grp <- scfa_long %>%
  filter(study_week == "week3",
         frequency == "2xdaily") 
var.test(x = once_grp$propionate_mean, 
         y = twice_grp$propionate_mean, 
         alternative = "two.sided") #p-value = 0.4277
# non-normal, equal variances
```

### pH
```{r}
# check sample size
pH_wkly %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>% 
  group_by(frequency, study_week) %>% 
  summarise(sample_size = n()) 

# normal distribution 
pH_df <- pH_wkly %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS")
shapiro.test(pH_df$ph_mean) #p-value = 0.001192
ggplot(pH_df, aes(x=ph_mean)) +
  geom_histogram() 
qqnorm(pH_df$ph_mean); qqline(pH_df$ph_mean)

# variances
once_grp <- pH_df %>%
  filter(study_week == "week3",
         frequency == "1xdaily") 
twice_grp <- pH_df %>%
  filter(study_week == "week3",
         frequency == "2xdaily") 
var.test(x = once_grp$ph_mean, 
         y = twice_grp$ph_mean, 
         alternative = "two.sided") #p-value = 0.0573

# non-normal, unequal varainces 
```


# Homework 5.4
Conduct the appropriate statistical tests to determine if patterns observed in plots are significant. Write your interpretations of the results as a comment after the statistical tests.
### Methane
```{r}
df1 <- breath_wkly %>% 
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(participant_id, study_week, ch4_mean) %>%
  spread(study_week, ch4_mean) %>%
  na.omit()
# non-normal, unequal variances, use wilcoxon test 
wilcox.test(x = df1$week1, 
            y = df1$week3, 
            alternative = "less", paired = TRUE, var.equal = FALSE)
```

### Acetate
```{r}
df2 <- scfa_long %>% 
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(participant_id, study_week, acetate_mean) %>%
  spread(study_week, acetate_mean) %>%
  na.omit()

# non-normal, equal variances, use wilcoxon test 
wilcox.test(x = df2$week1, 
            y = df2$week3, 
            alternative = "less", paired = TRUE)
```

### Propionate
```{r}
df3 <- scfa_long %>% 
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(participant_id, study_week, propionate_mean) %>%
  spread(study_week, propionate_mean) %>%
  na.omit()

# non-normal, equal variances, use wilcoxon test 
wilcox.test(x = df3$week1, 
            y = df3$week3, 
            alternative = "less", paired = TRUE)
```

### pH
```{r}
# non-normal, unequal variances, use wilcoxon test 
df4 <- pH_wkly %>% 
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>%
  select(participant_id, study_week, ph_mean) %>%
  spread(study_week, ph_mean) %>%
  na.omit()

# non-normal, equal variances, use wilcoxon test 
wilcox.test(x = df4$week1, 
            y = df4$week3, 
            alternative = "greater", paired = TRUE)
```

# ANOVA 
### Check assumptions
```{r}
# check normality of each group 
s7 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s7$delta_butyrate) #p-value = 0.6976

s8 <- scfa_wide %>%
  filter(supplement_consumed == "none") 
shapiro.test(s8$delta_butyrate) #p-value = 0.009887

s9 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s9$delta_butyrate) #p-value = 0.2783

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS") 
shapiro.test(s10$delta_butyrate) #p-value = 0.04406

s11 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize") 
shapiro.test(s11$delta_butyrate) #p-value = 0.06542

```

# Homework 5.5 
Repeat the processing of checking assumptions to conduct ANOVA on delta acetate and propionate. Write your interpretations of the results as a comment after the statistical tests.

### Acetate
##### plot 
```{r}
scfa_wide %>%
  ggplot(aes(x = supplement_consumed,
             y = delta_acetate,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Acetate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

##### normality 
```{r}
s1 <- scfa_wide %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$delta_acetate) #p-value = 0.674

s2 <- scfa_wide %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$delta_acetate) #p-value = 0.05369

s3 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$delta_acetate) #p-value = 0.8305

s4 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$delta_acetate) #p-value = 0.001526

s5 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$delta_acetate) #p-value = 0.3026

s6 <- scfa_wide %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$delta_acetate) #p-value = 0.9105

s7 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s7$delta_acetate) #p-value = 0.137

s8 <- scfa_wide %>%
  filter(supplement_consumed == "none") 
shapiro.test(s8$delta_acetate) #p-value = 0.3468

s9 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s9$delta_acetate) #p-value = 0.7713

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS") 
shapiro.test(s10$delta_acetate) #p-value = 0.9725

s11 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize") 
shapiro.test(s11$delta_acetate) #p-value = 0.6092

```

##### variances
```{r}
bartlett.test(delta_acetate ~ supplement_consumed, data = scfa_wide) #p-value = 1.167e-05
```

##### anova 
```{r}
# due to unequal variances we cannot use ANOVA 
aov_results <- aov(delta_acetate ~ supplement_consumed, data = scfa_wide)
summary(aov_results)
```


### Propionate

##### plot
```{r}
scfa_wide %>%
  ggplot(aes(x = supplement_consumed,
             y = delta_propionate,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Propionate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

##### normality 
```{r}
s1 <- scfa_wide %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$delta_propionate) #p-value = 6.507e-05

s2 <- scfa_wide %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$delta_propionate) #sample size too small 

s3 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$delta_propionate) #p-value = 0.2266

s4 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$delta_propionate) #p-value = 0.2669

s5 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$delta_propionate) #p-value = 0.4316

s6 <- scfa_wide %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$delta_propionate) #p-value = 0.8973

s7 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s7$delta_propionate) #p-value = 0.8389

s8 <- scfa_wide %>%
  filter(supplement_consumed == "none") 
shapiro.test(s8$delta_propionate) #p-value = 0.1025

s9 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s9$delta_propionate) #p-value = 0.1115

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS") 
shapiro.test(s10$delta_propionate) #p-value = 0.6636

s11 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize") 
shapiro.test(s11$delta_propionate) #p-value = 0.4711

```

##### variances
```{r}
bartlett.test(delta_propionate ~ supplement_consumed, data = scfa_wide) #p-value = 0.00628
```

##### anova 
```{r}
# due to unequal variances we cannot use ANOVA 
aov_results <- aov(delta_propionate ~ supplement_consumed, data = scfa_wide)
summary(aov_results)
```






