---
title: 'Bio 201: Lab 3 graphics'
author: "Kristi Gdanetz MacCready"
date: "9/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/Lab3/")
```

We always begin by loading the required packages. ggplot2 is included in the tidyverse package, and if not completed earlier, load cowplot. We will use cowplot in addition to ggplot when generating graphics because it simplifies the ggplot default themes to conform to figure standards set by most biological publications. ggplot2 is a plotting package that makes it simple to create complex plots from data in a data frame. It provides a more programmatic interface for specifying what variables to plot, how they are displayed, and general visual properties. Therefore, you only need minimal changes if the underlying data change or if you decide to change from a bar plot to a scatter plot. This helps in creating publication quality plots with minimal amounts of adjustments and tweaking.

The ggplot2 package or just “ggplot” as it is commonly known, is a powerful tool for generating figures. The gg in the name refers to the “Grammar of Graphics”, which is a way of thinking of figures as being a series of layers consisting. Originally described by Leland Wilkinson, the grammar has been updated and applied to R by Hadley Wickham, the package’s creator. According to the grammar, a figure consists of the underlying data, aesthetic mappings, geometric objects, scales, a coordinate system, statistical transformations, and facet specifications. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Import dataset

ggplot2 functions like data in the ‘long’ format, i.e., a column for every dimension, and a row for every observation. Well-structured data will save you lots of time when making figures with ggplot2. Tidying up the data is one of the biggest hurdles to learning R, so you will practice this for the next several weeks. 

### Homework 3.4
```{r}
# import curated data set, save as new data frame 
scfa_qc <- read_delim(file = "curated_data/SCFA_wkly.txt", 
                        delim = "\t", escape_double = FALSE, trim_ws = TRUE, na=c("NA"),
                        col_types = list())

# drop median columns

# keep only individuals who consumed full starch dose

# covert columns to snake case 

# add units to SCFA measurements 

# calculate total SCFA
```

# Plotting with ggplot 2
ggplot graphics are built step by step by adding new elements. Adding layers in this fashion allows for extensive flexibility and customization of plots.

To build a ggplot, you will use the following basic template that can be used for different types of plots:
```{r eval=FALSE, include=FALSE}
# don't run this
ggplot(data = <DATA>, 
       mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()
```

# Types of plots
Use the ggplot() function and bind the plot to a specific data frame using the "data" argument:
```{r include=FALSE}
ggplot(data = scfa_qc)
```

Define a mapping (using the aesthetic (aes) function), by selecting the variables to be plotted and specifying how to present them in the graph, e.g. as x/y positions or characteristics such as size, shape, color, etc.
```{r}
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg))
```

We set the axes, however we still have to tell R how we want the data values added. This is accomplished with ‘geoms’ – graphical representations of the data in the plot (points, lines, bars). ggplot2 offers many different geoms; you will use some common ones today, including:
```{r eval=FALSE, include=FALSE}
geom_point() #for scatter plots, dot plots, etc.
geom_boxplot() #for, well, boxplots!
geom_line() #for trend lines, time series, etc.  
geom_jitter() #for representation of all points in a group
geom_col() #for bar graphs, don't confuse with geom_bar()
```

To add a geom to the plot use the + operator. Because you have two continuous variables, let’s use geom_point() first: 
```{r}
ggplot(data = scfa_qc, #call dataset
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg)) + #assign data to axes
  	geom_point() #type of plot 
```

The + in the ggplot2 package is particularly useful because it allows you to modify existing ggplot objects. This means you can easily set up plot templates and conveniently explore different types of plots, so the above plot can also be generated with code like this:

```{r}
# Assign plot to a variable
plot1 <- ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg))
# Draw the plot
plot1 + geom_point()
```

Anything you put in the ggplot() function can be seen by any geom layers that you add (i.e., these are universal plot settings). This includes the x- and y-axis mapping called in aes().

You can also specify mappings for a given geom independently of the mappings defined globally in the ggplot() function.

The + sign used to add new layers must be placed at the end of the line containing the previous layer. If instead the + sign is added at the beginning of the line containing the new layer, ggplot2 will not add the new layer and will return an error message.

```{r eval=FALSE, include=FALSE}
# This is the correct syntax for adding layers
plot1 +
  	geom_point()
 
# This will not add the new layer and will return an error message
plot1 
  + geom_point()
```

# Customization + themes

Take a look at the ggplot2 cheat sheet, and think of ways you could improve the plot.
The axis names are informative, however they contain the underscores found in the column names. We can update these and add a title to the figure:
```{r}
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg, 
                     color = study_week)) +
    geom_point() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = "Individuals",
         y = "Fecal butyrate (mmol/kg)") 
```

Note that it is also possible to change the fonts of the plots. Remember we installed the cowplot package, this removes the default gray background and increases the default font sizes in figures automatically. In the end there is less code to write! 

After our manipulations, you may notice that the values on the x-axis are still not properly readable. Let’s change the orientation of the labels and adjust them vertically and horizontally to try to improve visibility. You can use a 90-degree angle, or experiment to find the appropriate angle for diagonally oriented labels:

```{r}
# flip x and y axes
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg, 
                     color = study_week)) +
    geom_point() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = "Individuals",
         y = "Fecal butyrate (mmol/kg)") +
    coord_flip()
```
Not much better, eh?

```{r}
# change angle of axis labels
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg, 
                     color = study_week)) +
    geom_point() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = "Individuals",
         y = "Fecal butyrate (mmol/kg)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5))
```
Rotating on the x-axis isn't really any better than swapping the x- and y- axes.

In reality the individual particpant IDs are not published in figures for research journals, so let's hide them: 
```{r}
# hide x-axis tick labels
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg, 
                     color = study_week)) +
    geom_point() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = "Individuals",
         y = "Fecal butyrate (mmol/kg)") +
    theme(axis.text.x = element_blank())
```
So the axes are cleaned up, but there are still so many data points here, it is impossible to draw any conclusions. These are some basic manipulations using a scatter plot as an example to improve the basic appearance of a plot. Our issue here is that we didn't choose an optimal plot type for our volume data, what if switch our type of plot?
```{r}
ggplot(data = scfa_qc, 
       mapping = aes(x = participant_id, 
                     y = butyrate_mean_mmol_kg, 
                     color = study_week)) +
    geom_line() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = "Individuals",
         y = "Fecal butyrate (mmol/kg)") +
    theme(axis.text.x = element_blank())
```
 ggplot can't make sense of this data structure for a line plot. We have to modifiy the data assigned to each axis, and as mentioned in the error message above, we need to set a group. 
```{r}
ggplot(data = scfa_qc, 
       mapping = aes(x = study_week, #switch from participant_id
                     y = butyrate_mean_mmol_kg, #remains the same
                     color = supplement_consumed, #switch from study_week
                     group = participant_id)) + #this is a new argument 
    geom_line() +
    labs(title = "Mean butyrate concentrations of individuals",
         x = NULL,
         y = "Fecal butyrate (mmol/kg)") 
```

As with the scatter plot, there are a lot of lines (each connecting an individual data point) here, and it is difficult to see patterns. One way we can deal with this is subsetting the data and creating plots with those subsets. Another strategy is summarising the data futher (e.g., calculate the mean for each supplement) and plot fewer values; we will cover that topic next week. 

# Arranging plots
In the section above you created a figure for just the butyrate data. The data frame contains measurements for two other SCFA, acetate and propionate, you also calculated the total SCFA. We would like to create figures for each of these, four figures total. 

### Generate individual plots
### Homework 3.5
Recreate the scatter plot from the above for each SCFA. Identify descriptive axis labels, make each plot a different color. Utilize copy and paste to reduce the amount of typing. Assign each figure to a new object. Website with some named ggplot colors: http://sape.inf.usi.ch/quick-reference/ggplot2/colour 
```{r}
plot_ace <- ggplot()
```

```{r}
plot_but <- ggplot()
```

```{r}
plot_pro <- ggplot()
```

```{r}
plot_tot <- ggplot()
```


### Combine plot panels 

We will use another function of cowplot to easily combine multiple figures. Cowplot has a lot of functions to organize panels of figures for publications, and some of these might be useful for your semester presentations, such as adding letter annotations to each panel of a figure:

```{r}
plot_grid(plot_ace, plot_but, plot_pro, plot_tot,
          labels = c("A", "B", "C", "D"),
          nrow = 2, ncol = 2)
```

# Faceting

In the section above you created multiple plots, and combined them into one larger multi-panel figure. These functions are useful if individual panels require a lot of customization, or are generated from different data frames. However if you have a tidy data set you can create a multi-panel figure with one function added to the ggplot call. The faceting function uses the provided argument, usually a factor (more on factors next week) which is a categorical group that describes one characteristic/measurement of each data point. Under the hood the data is subset by these factors, and an individual panel is created for each factor. 

### Without facets 
Subset the data frame for week 1 then week 3 of the study, plot the total SCFA measurements. Notice in the code below we pipe the filtered data frame directly to ggplot. We do not have to create additional dataframes, and use up a lot of memory.
```{r}
plot_wk1 <- scfa_qc %>%
  filter(study_week == "week1") %>%
  ggplot(aes(x = participant_id, 
             y = total_mmol_kg)) +
    geom_point(color = "black") +
    labs(title = "Week 1",
         x = "Individuals",
         y = "Total fecal SCFAs (mmol/kg)") +
    theme(axis.text.x = element_blank(),
          legend.position = "none")

plot_wk3 <- scfa_qc %>%
  filter(study_week == "week3") %>%
  ggplot(aes(x = participant_id, 
             y = total_mmol_kg)) +
    geom_point(color = "black") +
    labs(title = "Week 3",
         x = "Individuals",
         y = "Total fecal SCFAs (mmol/kg)") +
    theme(axis.text.x = element_blank(),
          legend.position = "none")

plot_grid(plot_wk1, plot_wk3, 
          nrow = 1, ncol = 2)
```

### With facets
We will use the code above as a template to create a scatter plot. We need to pass some additional arguments to aesthetics without assigning them to an axis: supplement_consumed, semester, study_week.

The new function added to the plot code here is “facet_wrap”. Open the help documentation and explore the default behavior and optional arguments.

```{r}
plot_f <- scfa_qc %>%
  filter(study_week == "week1" | study_week == "week3") %>%
  ggplot(aes(x = participant_id, 
             y = total_mmol_kg, 
             study_week, supplement_consumed, semester)) +
  geom_point(aes(color = study_week)) +
  facet_grid(~study_week) + 
  labs(x = "Individuals",
       y = "Total fecal SCFAs (mmol/kg)") +
  theme(axis.text.x = element_blank(),
        legend.position = "none")

plot_f
```

Complete any additional customizations to make the figure easy to read. Compare the facetted figure with the version you generated above where each panel was created manually. What differences in the layout of the figure do you notice? Write your answer in the bold block below:

> ** enter answer here ** 

### Homework 3.6
Plot the mean butyrate concentration for week 3 versus semester. In this plot, each facet should be one of the supplement types consumed during the study. Modify the plot settings so all labels are descriptive and easy to read. Tip: read the help documentation for facet_grid and modify some of the default arguments. 
```{r}
plot3_6 <- scfa_qc %>%
  filter() %>%
  ggplot()
```

# Exporting plots

As with other topics covered in this course, there are multiple ways to achieve the goal, including saving or exporting the graphics created in RStudio. For one-off figures use the mouse interactively to save the figure. However when completing a long analysis, especially with a dataset that is updated regularly (e.g., new data is added each semester to the undergrad cohort), it is the best practice to explicitly code the figure save/export. This allows you to save and compare versions of the same figures. In the directory for lab 3 create a folder called "figures". Save the plot you created in the facet section. 

```{r}
# preferred save method with cowplot
save_plot(filename = "figures/facet_plot.pdf",
          plot = plot_f,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

# option to save with ggplot functions 
ggplot2::ggsave(filename = "facet_plot.tiff",
                path = "figures/",
                device = "tiff",
                plot = plot_f,
                dpi = "retina")
```

### Homework 3.7
Export plot 3.6 
```{r}

```

-----
end