---
title: "Lab6 answers"
author: "Kristi Gdanetz MacCready"
date: "10/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Box Sync/Schmidt_Lab/Bio201_teaching/Protocol_Rmd_docs/")
```

# join

```{r}
#check number of rows
nrow(taxa_cat_curated) #55138
nrow(scfa_long) #1361

#check number of participants 
n_distinct(taxa_cat_curated$participant_id) #689
n_distinct(scfa_long$participant_id) #584

#identify shared columns = participant_id, study_week, semester
colnames(taxa_cat_curated) 
colnames(scfa_long)
```

```{r}
tax_scfa <- inner_join(scfa_long, taxa_cat_curated,
                 by = c("participant_id", "study_week", "semester")) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
View(tax_scfa)
```

```{r}
nrow(tax_scfa) #26878
n_distinct(tax_scfa$participant_id) #396
colnames(tax_scfa)
```

### Homework 6.1
* Write hypotheses to answer the question; do hydrogen breath levels differ between participants consuming fiber supplements? Create a plot of hydrogen levels:
```{r}
# hypothesis

# plot
breath_wkly %>%
  ggplot(aes(x = supplement_consumed, 
             y = h2_mean, 
             color = supplement_consumed)) + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Breath hydrogen (ppm)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

* Which assumptions need checked for the statistical test(s)? Check the assumptions in the space below:
```{r}
# sample size 

# normality

# variance
```

* Which statistical test(s) would you use to answer the question? Conduct the appropriate statistical test(s):
```{r}
# anova

# kruskall wallace 
```

* What is the answer to the question? Write your conclusion:
```{r}
# conclusion 
```

### Homework 6.2

Use the `tax_scfa` data frame, calculate the ratio of each SCFA in a new column (scfa/total). Check that your ratios sum to 1. 

```{r}
scfa_ratio <- tax_scfa %>%
  mutate(acetate_ratio = (acetate_mean / totalscfa_mean),
         butyrate_ratio = (butyrate_mean / totalscfa_mean),
         propionate_ratio = (propionate_mean / totalscfa_mean),
         ratio_check = (acetate_ratio + butyrate_ratio + propionate_ratio))
```

Calculate the ratios of Firmicutes (Fr = F / F+B) and Bacteroidetes (Br = B / F+B), and the overall ratio (Fr/Br).

```{r}
# filter for firmicutes
Fr <- scfa_ratio %>%
  filter(phylum == "Firmicutes") %>% 
  group_by(participant_id, study_week, frequency, 
           semester, supplement_consumed, quantity_compliant) %>%
  summarise(firmicutes_abund = sum(relative_abundance))

# filter for Bacteroidetes
Br <- scfa_ratio %>%
  filter(phylum == "Bacteroidetes <phylum>") %>% 
  group_by(participant_id, study_week, frequency, 
           semester, supplement_consumed, quantity_compliant) %>%
  summarise(bacteroidetes_abund = sum(relative_abundance)) 

# join two subsetted data frames
tax_ratio <- full_join(Fr, Br,
                       by = c("participant_id", "study_week", "frequency", 
                                "semester", "supplement_consumed", "quantity_compliant")) %>%
  # calculate ratios
  mutate(f_b_ratio = (firmicutes_abund / bacteroidetes_abund)) %>%
  # join with scfa_ratio data frame
  full_join(scfa_ratio, by = c("participant_id", "study_week", "frequency", 
                                "semester", "supplement_consumed", "quantity_compliant"))
```

### Homework 6.3
Determine if there is a relationship between the butyrate ratio and firmicutes (Fr). Conduct the statistical test and generate a plot. 
```{r}
# plot
tax_ratio %>%
  filter(phylum == "Firmicutes", #subset dataframe to firmicutes
         study_week == "week3", #subset for week3
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT"
         ) %>%
  ggplot(aes(x = butyrate_ratio,
             y = firmicutes_abund)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 mean butyrate (mmol/kg)") + 
  ylab("Firmicutes Abundance") 

# correlation test 
m63 <- tax_ratio %>%
  filter(phylum == "Firmicutes", #subset dataframe to firmicutes
         study_week == "week3", #subset for week3
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT") %>% 
  lm(butyrate_ratio ~ firmicutes_abund, data = .) #test relationship
summary(m63) #view results 
```

### Homework 6.4
Determine if there is a relationship between the propionate ratio and Bacteroidetes abundance. Conduct the statistical test and generate a plot. 
```{r}
# plot
tax_ratio %>%
  filter(phylum == "Bacteroidetes <phylum>", 
         study_week == "week3", #subset for week3
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT") %>%
  ggplot(aes(x = propionate_ratio,
             y = bacteroidetes_abund)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 propionate ratio (mmol/kg)") + 
  ylab("Bacteroidetes Abundance") 

# test
m64 <- tax_ratio %>%
  filter(phylum == "Bacteroidetes <phylum>", 
         study_week == "week3", #subset for week3
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT") %>% 
  lm(propionate_ratio ~ bacteroidetes_abund, data = .) #test relationship
summary(m64) #view results 
```

### Homework 6.5
Determine if there is a relationship between total SCFAs and acetate. Conduct the statistical test and generate a plot. 
```{r}
# plot
tax_ratio %>%
  ggplot(aes(x = acetate_mean,
             y = totalscfa_mean)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Weekly mean acetate (mmol/kg)") + 
  ylab("Weekly total SCFA (mmol/kg)") 

#Total SCFA vs. acetate
m65 <- tax_ratio %>%
  lm(acetate_mean ~ totalscfa_mean, data = .) 
summary(m65) 
```

### Homework 6.6
Import the curated pH+BSS data you generated during Lab4. Update the path and file name below as necessary.
```{r}
# import
pH_BSS_wkly <- read_delim(file = "~/Documents/UMich_Bio201_F19/Lab4/curated_data/pH_BSS_wkly.txt",
                          delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA"))
# join
tax_scfa_pH <- full_join(tax_ratio, pH_BSS_wkly,
           by = c("participant_id", "study_week")) %>%
  na.omit(semester)
```

Determine if there is a relationship between total SCFAs and pH Conduct the statistical test and generate a plot. 
```{r}
# plot
tax_scfa_pH %>%
  ggplot(aes(x = totalscfa_mean,
             y = pH_avg)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Weekly total SCFA (mmol/kg)") + 
  ylab("mean fecal pH") 

#Total SCFA vs. pH
m66 <- tax_scfa_pH %>%
  lm(totalscfa_mean ~ pH_avg, data = .) 
summary(m66) 
```

### Homework 6.7
Determine if there is a relationship between butyrate and pH Conduct the statistical test and generate a plot. 
```{r}
# plot
tax_scfa_pH %>%
  ggplot(aes(x = butyrate_mean,
             y = pH_avg)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Weekly mean butyrate (mmol/kg)") + 
  ylab("mean fecal pH") 

# Butyrate vs. pH
m67 <- tax_scfa_pH %>%
  lm(butyrate_mean ~ pH_avg, data = .) 
summary(m67)  
```






